---
title: "Iris for Botantists"
date: "10th January 2022"
output: 
  html_document: 
    code_folding: hide
#output: github_document
---

```{r setup, include=FALSE }
knitr::opts_chunk$set(echo = TRUE)
library(rgl)
library(reshape2)
library(tidyverse)

library(svglite)
#knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache.comments=FALSE, comment=NA, fig.fullwidth = TRUE, 
#  dev = "svglite", fig.ext = ".svg", autoprint = TRUE, echo = TRUE)
knitr::knit_hooks$set(webgl = hook_webgl, rgl = hook_rgl)
knitr.table.format = "html"

```

## Brief
*	Botanists require you to help them design an algorithm to identify flower types from four key measurements.
*	The data is a modified version of the iris data set.

[Original data](https://onlinelibrary.wiley.com/doi/10.1111/j.1469-1809.1936.tb02137.x "Link to journal")


## Exploratory Data Analysis
```{r,  message = FALSE, warning=FALSE}
#Original dataset
read.csv("https://gist.githubusercontent.com/curran/a08a1080b88344b0c8a7/raw/0e7a9b0a5d22642a06d3d5b9bcbad9890c8ee534/iris.csv") -> iris
#Modified local dataset
iris %>% colnames() -> org.col
read.csv("iris_data.csv") %>% setNames(org.col) -> iris



```
### Summary Statistics
*Iris setosa*

Three anomalies appear within *Iris setosa* subgroup. Sepal length contains a max value of 5.4 meters and two empty values for sepal width and petal length.
 
```{r,  message = FALSE, warning=FALSE}
knitr::kable(summary(iris[iris$species == "setosa",]))
```
For row 6 with 540 cm, the sepal width is 3.9 cm and petal length is 1.7 cm. 
```{r,  message = FALSE, warning=FALSE}
iris %>% filter(sepal_length == 540) %>% knitr::kable()
```
The mean sepal length of the four other occurrences of where either are present is 5.4 cm, which will replace the existing 540cm.
```{r,  message = FALSE, warning=FALSE}
iris %>% filter(sepal_width == 3.9 | petal_length == 1.7) %>% knitr::kable()
iris %>% filter(sepal_width == 3.9 | petal_length == 1.7) %>%
  mutate(sepal_length = case_when(sepal_length == 540 ~ NA_real_, TRUE ~ sepal_length)) %>%
    summarise(mean.sep.len = mean(sepal_length, na.rm = TRUE))  %>% knitr::kable()
```
Updated row 13.
```{r,  message = FALSE, warning=FALSE}
iris %>% mutate(sepal_length = case_when(sepal_length == 540 ~ 5.4, TRUE ~ sepal_length)) -> iris

```
The remaining issue relate to row 13 with missing values for sepal width and petal length. 
```{r,  message = FALSE, warning=FALSE}
iris %>% filter(is.na(sepal_width)|is.na(petal_length)) %>% knitr::kable()
```
Identify where sepal length is 4.8 cm and petal width is 0.1 cm and compute the mean based on flowers with the same measurement.
```{r,  message = FALSE, warning=FALSE}
iris %>% filter(sepal_length == 4.8 | petal_width == .1) %>% knitr::kable()


iris %>% filter(sepal_length == 4.8 | petal_width == .1) %>%
 
    summarise(mean.sep.wid = mean(sepal_width, na.rm = TRUE),
              mean.pet.len = mean(petal_length, na.rm = TRUE))  %>% knitr::kable()
 iris %>% mutate(sepal_width = case_when(is.na(sepal_width) ~ 3.3, TRUE ~ sepal_width)) %>%
         mutate(petal_length = case_when(is.na(petal_length) ~ 1.5, TRUE ~ petal_length)) -> iris
iris %>% filter(sepal_length == 4.8 & petal_width == .1) %>% knitr::kable()
```



Remaining species of *I. vericolor* and *I. virginica* have no issues.

*Iris versicolor*
```{r,  message = FALSE, warning=FALSE}
knitr::kable(summary(iris[iris$species == "versicolor",]))
```
*Iris virginica*
```{r,  message = FALSE, warning=FALSE}
knitr::kable(summary(iris[iris$species == "virginica",]))
```


### Data Visualisation

```{r,  message = FALSE, warning=FALSE}

iris %>% mutate(id = 1:n()) -> iris 

iris2 <- melt(iris, id.vars =  c("id", "species"))
iris2 %>% rename(cm = value) -> iris2


ggplot(iris2, aes(x = variable, y = cm, fill = variable)) +
  geom_boxplot() +
  facet_wrap(~species) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  

ggplot(iris2, aes(x = variable, y = cm, fill = species)) +
  geom_line( alpha = 0.1, aes( group = id, color = species)) +
geom_boxplot() 

   
```

## Models

### Cluster Analysis
*I. setosa* is differential but the other two species overlap preventing identification.

```{r,  message = FALSE,warning=FALSE}
ggplot(data=iris, aes(x=sepal_length, y=petal_length, color=species)) + geom_point(aes( fill = species),size =3, alpha = 1, shape = 16) 
ggplot(data=iris, aes(x=sepal_length, y=petal_width, color=species)) +  geom_point(aes( fill = species),size =3, alpha = 1, shape = 16)  
ggplot(data=iris, aes(x=petal_length, y=petal_width, color=species)) +  geom_point(aes( fill = species),size =3, alpha = 1, shape = 16)
ggplot(data=iris, aes(x=petal_length, y=petal_width, color=species)) +  geom_point(aes( fill = species),size =3, alpha = 1, shape = 16) +
  labs(x ="Petal length (cm)", y = "Petal width (cm)") +
  #geom_vline(xintercept=2.5, linetype="dashed", color = "blue") +
  geom_segment(aes(x = 2.5, y = 1.6, xend = 8, yend = 1.6),  linetype="dashed", color = "#C77CFF") +
  geom_segment(aes(x = 0.8, y = 0.7, xend = 4.9, yend = 0.7),  linetype="dashed", color = "#C77CFF") +

  geom_segment(aes(x = 4.9, y = 0, xend = 4.9, yend = 2.5),  linetype="dashed", color = "#00BFC4") +
  geom_segment(aes(x = 2.5, y = 0, xend = 2.5, yend = 2.5),  linetype="dashed", color = "#00BFC4") 



```
In three dimensions
```{r,   webgl = TRUE, echo=FALSE,warning=FALSE}
options(rgl.useNULL=TRUE)
library(rgl)
rgl::setupKnitr()
#install.packages("rgl")
#library(rglwidget)


colors <- c("#F8766D", "#00BA38", "#619CFF")
iris$color <- colors[ as.numeric( as.factor(iris$species) ) ]

# Plot
plot3d( 
  x=iris$petal_length, y=iris$petal_width, z=iris$sepal_length, 
  col = iris$color, 
  type = 's', 
  radius = .1,
  xlab="Sepal Length (cm)", ylab="Petal Width (cm)", zlab="Petal Length (cm)") 
#-> iris.3d
           
```
### Logistic Regression
```{r,  message = FALSE,warning=FALSE}

iris2 %>% filter(species != 'setosa') %>% 
    mutate(prob = ifelse(species == "versicolor", 1, 0)) -> iris3


 probX = function(p, model) {
  data.frame(prob=p, 
             xval = (qnorm(p) - coef(model)[1])/coef(model)[2])
}

 
log <- glm(prob ~ cm, data=iris3[iris3$variable == 'petal_width',], family=binomial(link="logit"))
petal.w = probX(c(0.5), log)
petal.w$variable <-"petal_width"

log <- glm(prob ~ cm, data=iris3[iris3$variable == 'petal_length',], family=binomial(link="logit"))
petal.l = probX(0.5, log)
petal.l$variable <-"petal_length"


log <- glm(prob ~ cm, data=iris3[iris3$variable == 'sepal_length',], family=binomial(link="logit"))
sepal.l = probX(0.5, log)
sepal.l$variable <-"sepal_length"

log <- glm(prob ~ cm, data=iris3[iris3$variable == 'sepal_width',], family=binomial(link="logit"))
sepal.w = probX(0.5, log)
sepal.w$variable <-"sepal_width"

petal.w %>% bind_rows(petal.l, sepal.l, sepal.w) -> data.inter
data.inter %>% knitr::kable()
rm(petal.l, petal.w, sepal.w, sepal.l,log)
iris3 %>%
  filter(variable == "petal_width" ) %>%
  group_by(prob) %>%
  mutate(breaks = cut(cm, breaks=seq(0.1,2.5,0.2), labels=seq(0.2,2.5,0.2), 
                      include.lowest=TRUE),
         breaks = as.numeric(as.character(breaks))) %>%
  group_by(prob, breaks) %>% 
  summarise(n = n()) %>%
  mutate(pct = ifelse(prob==0, n/sum(n), 1 - n/sum(n))) %>% 
  mutate(variable = "petal_width")-> data.hist.p.w.

iris3 %>%
  filter(variable == "petal_length" ) %>%
  group_by(prob) %>%
  mutate(breaks = cut(cm, breaks=seq(1,6.9,0.2), 
                      labels=seq(1.25,6.9,0.2), 
                      include.lowest=TRUE),
         breaks = as.numeric(as.character(breaks))) %>%
  group_by(prob, breaks) %>% 
  summarise(n = n()) %>%
  mutate(pct = ifelse(prob==0, n/sum(n), 1 - n/sum(n))) %>% 
  mutate(variable = "petal_length")-> data.hist.p.l.

iris3 %>%
  filter(variable == "sepal_width" ) %>%
  group_by(prob) %>%
  mutate(breaks = cut(cm, breaks=seq(2,4.4,0.2), labels=seq(2.2,4.4,0.2), 
                      include.lowest=TRUE),
         breaks = as.numeric(as.character(breaks))) %>%
  group_by(prob, breaks) %>% 
  summarise(n = n()) %>%
  mutate(pct = ifelse(prob==0, n/sum(n), 1 - n/sum(n))) %>% 
  mutate(variable = "sepal_width")-> data.hist.s.w.

iris3 %>%
  filter(variable == "sepal_length" ) %>%
  group_by(prob) %>%
  mutate(breaks = cut(cm, breaks=seq(4.3,7.9,0.2), 
                      labels=seq(4.4,7.9,0.2), 
                      include.lowest=TRUE),
         breaks = as.numeric(as.character(breaks))) %>%
  group_by(prob, breaks) %>% 
  summarise(n = n()) %>%
  mutate(pct = ifelse(prob==0, n/sum(n), 1 - n/sum(n))) %>% 
  mutate(variable = "sepal_length")-> data.hist.s.l.



  ggplot( ) + 
geom_point(data  = iris3, aes(x=cm, y=prob, group = variable,fill= variable,  color = variable), alpha=.5) +
  stat_smooth(data  = iris3,aes(x=cm, y=prob, group = variable,fill= variable,  color = variable),method="glm", 
              method.args=list(family=binomial(link="probit"))) +
   geom_segment(data = data.inter, aes(x=xval, xend=xval, y=0, yend=prob), colour="#619CFF",linetype = "dashed") +
 # geom_segment(data=d, aes(x=rng[1], xend=xval, y=prob, yend=prob), colour="red") +
  geom_text(data = data.inter, aes(label=round(xval, 1), x=xval, y=-0.03), size=3, colour="#619CFF") +
      geom_segment(data=data.hist.p.w., size=4, show.legend=FALSE,
                aes(x=breaks, xend=breaks, y=prob, yend=pct, color = variable), alpha = 0.3) +
         geom_segment(data=data.hist.p.l., size=4, show.legend=FALSE,
               aes(x=breaks, xend=breaks, y=prob, yend=pct, color = variable), alpha = 0.3) +
      geom_segment(data=data.hist.s.w., size=4, show.legend=FALSE,
                aes(x=breaks, xend=breaks, y=prob, yend=pct, color = variable), alpha = 0.3) +
         geom_segment(data=data.hist.s.l., size=4, show.legend=FALSE,
               aes(x=breaks, xend=breaks, y=prob, yend=pct, color = variable), alpha = 0.3) +
   #  scale_y_continuous(breaks = 1:0,labels=c("I. versicolor","I. virginica")) #+
     scale_y_continuous("Probability",breaks = c(1,0.75,0.5,0.25,0),labels=c("I. versicolor","0.75","0.5","0.25","I. virginica")) #+

iris %>% filter(species != 'setosa') %>%   mutate(prob = ifelse(species == "versicolor", 1, 0)) -> iris4
model <- glm (prob ~ .-id-species-color, data = iris4, family = binomial)
model
summary(model)
plot(model)

predict <- predict(model, type = 'response')
#confusion matrix
table(iris4$prob, predict > 0.5)
predict(model, iris4, type="response") -> results
table(results > 0.5)
model <- glm (prob ~ .-id-species-color, data = iris4, family = binomial(logit))
```


### Decision Trees
C50 Algorithm
```{r,  message = FALSE,warning=FALSE}
library(tree)

tree(formula = as.factor(species) ~ petal_length + petal_width , data = iris) -> tree1
#Number of terminal nodes:  5 
plot(tree1)
text(tree1)
pred = predict(tree1, type="class")
knitr::kable(table(iris$species,pred))

plot(iris$petal_length,iris$petal_width,pch=19,col=as.numeric(as.factor(iris$species)))
partition.tree(tree1,label="Species",add=TRUE)
legend(1,2.5,legend=unique(as.factor(iris$species)),col=unique(as.numeric(as.factor(iris$species))),pch=19)


tree(formula = as.factor(species) ~ petal_length + petal_width + sepal_length +sepal_width , data = iris) -> tree3
#Number of terminal nodes:  5 
plot(tree3)
text(tree3)
pred = predict(tree3, type="class")
knitr::kable(table(iris$species,pred))
table(iris$species,pred) -> confMat
sum(diag(confMat))/sum(confMat)


library('C50') # load the package
irTree <- C5.0(iris[,-5:-7], as.factor(iris$species ))
summary(irTree) # view the model components  

plot(irTree, main = 'Iris decision tree') # view the model graphically  


irTree <- C5.0(iris[,-5:-7], as.factor(iris$species ), rules = TRUE)
summary(irTree) # view the model components  
 # view the model graphically  


```
Classification and Regression Tree (CART)

```{r,  message = FALSE,warning=FALSE}

library(rpart)
#install.packages("rattle")
library(rattle)

tree4 <- rpart(species ~ petal_length + petal_width + sepal_length +sepal_width, data=iris, method = "class")
fancyRpartPlot(tree4, main="Iris")

pred = predict(tree4, type="class")
confMat <- table(iris$species,pred)
knitr::kable(confMat)
sum(diag(confMat))/sum(confMat)

# plot decision tree

```
```{r,  message = FALSE,warning=FALSE, include = FALSE}

iris %>% filter( petal_length < 2.5) %>% group_by(species) %>% count()
iris %>% filter( petal_length > 4.9 | petal_width > 1.6) %>% group_by(species) %>% count()
iris %>% filter( petal_length <= 4.9 | petal_width <= 1.6) %>% group_by(species) %>% count()

iris %>% filter(petal_width >= 1.8) %>% group_by(species) %>% count()
iris %>% filter(petal_width >= 1.75 | petal_length >= 4.95) %>% group_by(species) %>% count()

iris %>% filter( petal_width < 1.6 )  %>% filter( petal_length < 4.9 ) %>% group_by(species) %>% count()
iris %>% filter( petal_width > 1.6 )  %>% filter( petal_length > 4.9 ) %>% group_by(species) %>% count()

iris %>% filter( petal_length <= 4.9 ) %>% group_by(species) %>% count()
iris %>% filter(petal_width < 1.8) %>% group_by(species) %>% count()
iris %>% filter(petal_width < 1.75 | petal_length < 4.95) %>% group_by(species) %>% count()

iris %>% filter(petal_width < 1.631638 | petal_length < 	4.863465) %>% group_by(species) %>% count()
iris %>% filter( petal_length <= 4.9 | petal_width <= 1.6) %>% group_by(species) %>% count()

```




In three dimensions for *I. virsicolor* and *I. virginica*.
```{r,   webgl = TRUE, echo=FALSE,warning=FALSE}
options(rgl.useNULL=TRUE)
library(rgl)
rgl::setupKnitr()
#install.packages("rgl")
#library(rglwidget)


colors <- c( "#00BA38", "#619CFF")
iris4$color <- colors[ as.numeric( as.factor(iris4$species) ) ]

# Plot
plot3d( 
  x=iris4$petal_length, y=iris4$petal_width, z=iris4$sepal_length, 
  col = iris$color, 
  type = 's', 
  radius = .1,
  xlab="Sepal Length (cm)", ylab="Petal Width (cm)", zlab="Petal Length (cm)") 
```