---
title: "Iris for Botantists"
date: "10th January 2022"
#output: html_document
output: github_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(rgl)

knitr::knit_hooks$set(webgl = hook_webgl)
```

## Brief

*	Botanists require you to help them design an algorithm to identify flower types from 4 key measurements.
*	The data is stored in iris.csv and is on your loop page.
*	Give some univariate statistics to describe the data.
*	What should you do with the missing data?
Correlations
*	Can you come up with a rules based algorithm to help the botanist?
Field
*	Teams of 2 people
Send me by email a 7-8 page ppt slide show presentation. It should contain your assumptions, analysis and conclusions. Try and get to the point quickly and only display the relevant information. 
I am looking for people to make an attempt at the analysis and marks will be awarded for effort and analytical thought. Don’t be afraid to have a go and demonstrate whatever knowledge you currently have. 
Projects can be completed in groups of a maximum of 3. 

https://www.jstor.org/stable/2394164?seq=1#metadata_info_tab_contents
http://www.dicook.org/files/jsm19/slides#45


## Explortatry Analysis
```{r}
read.csv("https://gist.githubusercontent.com/curran/a08a1080b88344b0c8a7/raw/0e7a9b0a5d22642a06d3d5b9bcbad9890c8ee534/iris.csv") -> iris

summary(iris)
library(reshape2)
library(tidyverse)
iris2 <- melt(iris, id.vars =   "species")
iris2 %>% rename(cm = value) -> iris2


ggplot(iris2, aes(x = variable, y = cm, fill = variable)) +
  geom_boxplot() +
  facet_wrap(~species) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
  

library(ggplot2)
library(GGally)

        ggpairs(iris,
            columns = c(1:4),
            title="Paired Plots of Iris ", 
            upper = list( combo = "facethist"),
            lower = list(combo =  'dot_no_facet',   alpha=0.2), #regrees
            diag = list(discrete = "barDiag", continuous="densityDiag"),
            mapping=ggplot2::aes(colour = species, fill=species, alpha=0.1))    
```
Take a closer look at sepal length, petal length and petal width.
```{r}
ggplot(data=iris, aes(x=sepal_length, y=petal_length, color=species)) + geom_point(aes(size = petal_width)) + geom_smooth(method="lm")
ggplot(data=iris, aes(x=sepal_length, y=petal_width, color=species)) + geom_point(size=3) + geom_smooth(method="lm")  
ggplot(data=iris, aes(x=petal_length, y=petal_width, color=species)) + geom_point(size=3) + geom_smooth(method="lm")


```
In three dimensions
```{r}


colors <- c("red", "green", "blue")
iris$color <- colors[ as.numeric( as.factor(iris$species) ) ]

# Plot
plot3d( 
  x=iris$petal_length, y=iris$petal_width, z=iris$sepal_length, 
  col = iris$color, 
  type = 's', 
  radius = .1,
  xlab="Sepal Length", ylab="Sepal Width", zlab="Petal Length")

# To display in an R Markdown document:
# rglwidget()

# To save to a file:
rglwidget()
                        
```

